Visual narratives of the Covid-19 pandemic

### 2022-09-18 JDSSV_manuscript_submitted.tex
Updates - final round of revisions?

\begin{figure*}
	\includegraphics[width = 0.98\textwidth]{Figures_Web/who_totalcases_choro.png}
	\caption{Choropleth map of the Covid-19 cases world-wide by country. Source: WHO \url{https://covid19.who.int}}
	\label{fig:choro1}
\end{figure*}

@@

\begin{figure*}
	\includegraphics[width = 0.98\textwidth]{Figures_Web/wp_totaldeaths_propsymb.png}
	\caption{Proportional symbol map showing the cumulated reported Covid-19 related deaths adjusted for population world-wide by country. Source: Washington Post \url{https://www.washingtonpost.com/graphics/2020/world/mapping-spread-new-coronavirus/}}
	\label{fig:propsymb}
\end{figure*}

@@

The narrative of the global perspective seems to be highly limited to the aspect of a pandemic affecting the entire globe: none of the above viusalisations intends to provide a deeper insight into the spatial distribution of the phenomenon, neither within the administratively motivated spatial borders nor across them. 
The use of maps to assess the global pandemic more often focuses on the comparative aspects: how are we doing as opposed to our neighbors? Which strategy to fight Covid-19 is better? 
We will examine these topics more closely in Section~\ref{sec:rankings}.

\begin{figure}
\centering
\begin{subfigure}[c]{.45\textwidth}
\includegraphics[width=\textwidth]{covid-tracking-hex-cartogram}
\caption{Hexagonal Cartogram of Covid-19 cases in the United States by state. The COVID Tracking Project \url{https://covidtracking.com/data/charts/us-daily-positive-cases}.}\label{fig:hex-cartogram}
\end{subfigure}\hfill
\begin{subfigure}[c]{.45\textwidth}
\includegraphics[width=\textwidth]{china-dotmap}
\caption{A dot-density map of COVID cases in China \citep{fieldMappingCoronavirusResponsibly}.}\label{fig:china-dot-map}
\end{subfigure}
% \begin{subfigure}[t]{.45\textwidth}
% \includegraphics[width=\textwidth]{nyt-covid-scaled-choro}
% \caption{Choropleth Map of covid cases per country, scaled by population. Source: New York Times \url{https://www.nytimes.com/interactive/2021/world/covid-cases.html}.}\label{fig:scaled-choro}
% \end{subfigure}
\caption{Modifications of choropleth maps which attempt to address the area bias of choropleths in different ways. Both methods are applied at the national scale rather than at global scale because of limitations with the visual form or data quality.}
\end{figure}

@@

Visualisations are typically created for a specific purpose, and during a pandemic it is particularly critical to show the evolution of the situation over time, as there are new developments on a daily basis. 
Typically, this means that cases, hospitalizations, and/or deaths are plotted on the y-axis, with date shown on the x-axis, as in \Cref{fig-jbm-tweet} and \Cref{fig:wapo-covid-time-series}. 
This not only allows individuals to compare the present to past case counts, but also allows for forecasting and prediction of the future based on the current state. 
Time series plots are the most natural choice for these goals, and over the course of the pandemic, many different attempts have been made to show time series information using different graphical forms.

\begin{figure}
\centering
\includegraphics[width=.99\linewidth]{wapo-covid-monitoring}
\caption{COVID monitoring summary charts on the Washington Post's Coronavirus page. Source: \url{https://www.washingtonpost.com/coronavirus/}, 2022-01-17.}
\label{fig:wapo-covid-time-series}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=.5\linewidth]{wapo-covid-context}
\caption{Some outlets provide contextual information to assist readers with interpreting time-series plots. Source: Washington Post Coronavirus Tracking, \url{https://www.washingtonpost.com/graphics/2020/national/coronavirus-us-cases-deaths/}.}
\label{fig:wapo-context}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=.75\linewidth]{frequency-domain}
\caption{Frequency-domain case count graphics. Here, each 1000 deaths are represented by a single line segment whose slope represents the time taken to reach the next 1000 deaths. This provides viewers with a sense of the pace of the epidemic, rather than the raw case counts in more standard time-series representations. This representation can also be shown in the auditory domain, providing access to those who are typically excluded from visualizations due to vision loss. Source: Romain Vuillemot, \url{https://observablehq.com/@romsson/sonification-of-covid-19-data}.}
\label{fig:freq-domain}
\end{figure}

@@

\begin{figure}\centering
\includegraphics[width=.45\linewidth]{nyt-spiral}\hfill\includegraphics[width=.45\linewidth]{spiral-rework}
\caption{Covid cases in the United States, 2020-2022. The original graph from the New York Times is on the left, and a reenvisioned version which is more perceptually friendly is on the right. Source: (Left) New York Times \url{https://www.nytimes.com/2022/01/06/opinion/omicron-covid-us.html}, (Right) Covid Spiral, by Sourya Shrestha, \url{https://github.com/souryashrestha/covid_spiral}.}
\label{fig:nyt-spiral}
\end{figure}

Other designers used the capabilities of web graphics to show the time component in the data through animation. In some cases, this took the form of animated bubble plots XXX add XXX, while others dispensed with the traditional form of a line graph altogether, showing rising case counts using a "bar chart race": an animated series of bar charts over time that show how case counts in each country are increasing. A typical example can be found on YouTube at \url{https://www.youtube.com/watch?v=8WH-bTH6GqI}; a screenshot is shown in \Cref{fig:barchart-race}. This type of chart is effective in showing instantaneous changes in counts over days for countries with an extremely high number of cases, but may make it difficult for the viewer to process how cases are changing in countries that do not dominate case counts. By focusing viewer attention on changes in relative totals of case counts, the graphic tends to hide steady, proportional increases in cases across the globe, making it somewhat difficult to perceive large global trends relative to small changes in comparative case counts.

\begin{figure}
\centering
\includegraphics[width=.8\linewidth]{Figures_Web/Bar-chart-race-screenshot}
\caption{A screenshot from Stats on Clock's bar chart race showing counts of covid cases over time throughout the pandemic. Source: \url{https://www.youtube.com/watch?v=8WH-bTH6GqI}}\label{fig:barchart-race}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=.8\linewidth]{Figures_Web/BBC_flower_time_series}
\caption{The BBC Visual and Data Journalism Team's Covid flower, showing the cases and deaths over time. Source: \url{https://www.bbc.co.uk/news/resources/idt-7464500a-6368-4029-aa41-ab94e0ee09fb}}\label{fig:bbc-flower}
\end{figure}
% BBC flowers w/ sonification https://www.bbc.co.uk/news/resources/idt-7464500a-6368-4029-aa41-ab94e0ee09fb

@@

\begin{figure}\centering
\includegraphics[width=\linewidth]{Figures_Web/ft_deathrates_bar}\\
\includegraphics[width=\linewidth]{Figures_Web/ft_deathcounts_area}
\caption{Ordered death rates in countries and the temporal evolution of the death toll share in major world regions. Source: Financial Times, \url{https://www.ft.com/content/a2901ce8-5eb7-4633-b89c-cbdf5b386938}}
\label{fig:ft-death-ranking}
\end{figure}

@@

One issue with assessing the use of log scales is that their effectiveness changes with the stage of the pandemic and the amount and variety of data shown. Initially, log scales were incredibly useful at showing case counts, because minimal mitigation measures were in place and the growth of case counts (or presumptive positive cases, in absence of available testing) was fairly close to exponential. In addition, the use of log scales allowed for the comparison of nominal cases across entities with large population differences: in the US, we could compare cases in New York and California with cases in Michigan and Washington, even though the population of Michigan and Washington are much lower than the population of either New York or California.

@@

\caption[One problem with log scales is that if there is a background level of spread, it can be hard to notice the introduction of an additional source of exponential spread]{One problem with log scales is that if there is a background level of spread, it can be hard to notice the introduction of an additional source of exponential spread. Linear scales do not have this problem - the exponential source is noticeable very quickly in the total line, but on the log scale it is much harder to discern when the exponential source causes the total line to diverge from the background. In the top-right corner, it is difficult to identify that there is an exponential increase in cases amid the baseline, even though the exponential source makes up approximately 50\% of the cases at the end of the time period shown.}\label{fig:log-scale-failures}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=.8\linewidth]{ft-reference-lines}
\caption{An annotated time series from the Financial Times which appeared on Twitter on November 12, 2020. This chart has annotations which show the decreasing $R_{eff}$ in successive peaks of the pandemic, with periods of $R_{eff}<1$ between surges in cases. These annotations assist the reader with drawing complex conclusions from the chart, but are difficult to automate and thus tend to be manually curated. Source: Mark Gubrud (@mgubrud), \url{https://twitter.com/mgubrud/status/1326784082399944704}.}
\label{fig:log-scale-ref-multiple}
\end{figure}

@@

\section{Summary and discussion} \label{sec:summary}
Throughout the pandemic (thus far), visual representations of data have been an integral part of scientific communication. While not always optimal, these graphics have attempted to provide meaningful context, to encourage individual and collective action, and to help individuals grapple with the scale of the pandemic in cases, deaths, vaccinations, and interventions. As in any developing situation, choices made at some points in the pandemic did not always persist - many news outlets refined their charts and approach to the design of graphics over the course of the pandemic. This evolution of graphical forms has provided us with an opportunity to evaluate what worked and did not -- and why -- in a context of broad general interest.

@@

@@

First ideas have been presented at DSSV 2020. We thank our colleagues and the editors for valuable feedback and encouragement to complete this manuscript.

>>>>

\begin{figure*}
	\includegraphics[width = 0.98\textwidth]{Figures_Web/who_totalcases_choro.png}
	\caption{Choropleth map of the Covid-19 cases world-wide by country \citep{WHOCoronavirusCOVID19}.}
	\label{fig:choro1}
\end{figure*}

@@

\begin{figure*}
	\includegraphics[width = 0.98\textwidth]{Figures_Web/wp_totaldeaths_propsymb.png}
	\caption{Proportional symbol map showing the cumulated reported Covid-19 related deaths adjusted for population world-wide by country \citep{alcantaraMappingWorldwideSpread}.}
	\label{fig:propsymb}
\end{figure*}

@@

The narrative of the global perspective seems to be highly limited to the aspect of a pandemic affecting the entire globe: none of the above viusalisations intends to provide a deeper insight into the spatial distribution of the phenomenon, neither within the administratively motivated spatial borders nor across them. 
The use of maps to assess the global pandemic more often focuses on the comparative aspects: how are we doing as opposed to our neighbors? Which strategy to fight Covid-19 is better? 
We will examine these topics more closely in \Cref{sec:rankings}.

\begin{figure}
\centering
\begin{subfigure}[c]{\textwidth}
\centering
\includegraphics[width=.65\textwidth]{covid-tracking-hex-cartogram}
\caption{Hexagonal Cartogram of Covid-19 cases in the United States by state \citep{StateCases1M2021}.}\label{fig:hex-cartogram}
\end{subfigure}

\begin{subfigure}[c]{\textwidth}
\centering
\includegraphics[width=\textwidth]{china-dotmap}
\caption{A dot-density map of COVID cases in China \citep{fieldMappingCoronavirusResponsibly}.}\label{fig:china-dot-map}
\end{subfigure}
% \begin{subfigure}[t]{.45\textwidth}
% \includegraphics[width=\textwidth]{nyt-covid-scaled-choro}
% \caption{Choropleth Map of covid cases per country, scaled by population. Source: New York Times \url{https://www.nytimes.com/interactive/2021/world/covid-cases.html}.}\label{fig:scaled-choro}
% \end{subfigure}
\caption{Modifications of choropleth maps which attempt to address the area bias of choropleths in different ways. Both methods are applied at the national scale rather than at global scale because of limitations with the visual form or data quality.}
\end{figure}
% 
% \begin{minipage}{\linewidth}
% \begin{minipage}{.7\linewidth}
% \includegraphics[width=\textwidth]{covid-tracking-hex-cartogram}
% \end{minipage}
% \begin{minipage}{.29\linewidth}
% \captionof{figure}{Hexagonal Cartogram of Covid-19 cases in the United States by state. The COVID Tracking Project \url{https://covidtracking.com/data/charts/us-daily-positive-cases}.}\label{fig:hex-cartogram}
% \end{minipage}
% \end{minipage}
% 
% \begin{minipage}{\linewidth}
% \begin{minipage}{.7\linewidth}
% \includegraphics[width=\textwidth]{china-dotmap}
% \end{minipage}
% \begin{minipage}{.29\linewidth}
% \captionof{figure}{A dot-density map of COVID cases in China \citep{fieldMappingCoronavirusResponsibly}.}\label{fig:china-dot-map}
% \end{minipage}
% \end{minipage}

@@

Visualisations are typically created for a specific purpose, and during a pandemic it is particularly critical to show the evolution of the situation over time, as there are new developments on a daily basis. 
Typically, this means that cases, hospitalizations, and/or deaths are plotted on the y-axis, with date shown on the x-axis, as in \Cref{fig:jbm-tweet} and \Cref{fig:wapo-covid-time-series}. 
This not only allows individuals to compare the present to past case counts, but also allows for forecasting and prediction of the future based on the current state. 
Time series plots are the most natural choice for these goals, and over the course of the pandemic, many different attempts have been made to show time series information using different graphical forms.

\begin{figure}
\centering
\includegraphics[width=.99\linewidth]{wapo-covid-monitoring}
\caption{COVID monitoring summary charts on the Washington Post's Coronavirus page \citep{Coronavirus2022}.}
\label{fig:wapo-covid-time-series}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=.5\linewidth]{wapo-covid-context}
\caption{Some outlets provide contextual information to assist readers with interpreting time-series plots \citep{CoronavirusCasesTracking2022}.}
\label{fig:wapo-context}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=.75\linewidth]{frequency-domain}
\caption{Frequency-domain case count graphics. Here, each 1000 deaths are represented by a single line segment whose slope represents the time taken to reach the next 1000 deaths. This provides viewers with a sense of the pace of the epidemic, rather than the raw case counts in more standard time-series representations. This representation can also be shown in the auditory domain, providing access to those who are typically excluded from visualizations due to vision loss \citep{vuillemotSonificationCOVID19Data2020}.}
\label{fig:freq-domain}
\end{figure}

@@

\begin{figure}\centering
\includegraphics[width=.45\linewidth]{nyt-spiral}\hfill\includegraphics[width=.45\linewidth]{spiral-rework}
\caption{Covid cases in the United States, 2020-2022. The original graph from the New York Times is on the left \citep{shamanOpinionHereWhen2022}, and a reenvisioned version which is more perceptually friendly is on the right \citep{shresthaCovidSpiral2022}.}
\label{fig:nyt-spiral}
\end{figure}

Other designers used the capabilities of web graphics to show the time component in the data through animation. In some cases, this took the form of animated bubble plots XXX add XXX, while others dispensed with the traditional form of a line graph altogether, showing rising case counts using a "bar chart race": an animated series of bar charts over time that show how case counts in each country are increasing. A typical example can be found on YouTube\citep{clockCoronavirusCasesBar2022}; a screenshot is shown in \Cref{fig:barchart-race}. This type of chart is effective in showing instantaneous changes in counts over days for countries with an extremely high number of cases, but may make it difficult for the viewer to process how cases are changing in countries that do not dominate case counts. By focusing viewer attention on changes in relative totals of case counts, the graphic tends to hide steady, proportional increases in cases across the globe, making it somewhat difficult to perceive large global trends relative to small changes in comparative case counts.

\begin{figure}
\centering
\includegraphics[width=.8\linewidth]{Figures_Web/Bar-chart-race-screenshot}
\caption{A screenshot from Stats on Clock's bar chart race showing counts of covid cases over time throughout the pandemic \citep{clockCoronavirusCasesBar2022}.}\label{fig:barchart-race}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=.8\linewidth]{Figures_Web/BBC_flower_time_series}
\caption{The BBC Visual and Data Journalism Team's Covid flower, showing the cases and deaths over time \citep{thebbcvisualanddatajournalismteamCoronavirusHowCan2020}.}\label{fig:bbc-flower}
\end{figure}
% BBC flowers w/ sonification https://www.bbc.co.uk/news/resources/idt-7464500a-6368-4029-aa41-ab94e0ee09fb

@@

\begin{figure}\centering
\includegraphics[width=\linewidth]{Figures_Web/ft_deathrates_bar}\\
\includegraphics[width=\linewidth]{Figures_Web/ft_deathcounts_area}
\caption{Ordered death rates in countries and the temporal evolution of the death toll share in major world regions \citep{CoronavirusTrackerLatest2021}.}
\label{fig:ft-death-ranking}
\end{figure}

@@

One issue with assessing the use of log scales is that their effectiveness changes with the stage of the pandemic and the amount and variety of data shown. Initially, log scales were incredibly useful at showing case counts, because minimal mitigation measures were in place and the growth of case counts (or presumptive positive cases, in absence of available testing) was fairly close to exponential. In addition, the use of log scales allowed for the comparison of total cases across entities with large population differences: in the US, we could compare cases in New York and California with cases in Michigan and Washington, even though the populations (and corresponding case counts) of Michigan and Washington are much lower than the populations (and case counts) of either New York or California.

@@

\caption[One problem with log scales is that if there is a background level of spread, it can be hard to notice the introduction of an additional source of exponential spread]{One problem with log scales is that if there is a background level of spread, it can be hard to notice the introduction of an additional source of exponential spread. Linear scales do not have this problem - the exponential source is noticeable very quickly in the total line, but on the log scale it is much harder to discern when the exponential source causes the total line to diverge from the background. In the top-right corner, it is difficult to identify that there is an exponential increase in cases amid the baseline of a uniform level of spread over time, even though the exponential source makes up approximately 50\% of the cases at the end of the time period shown.}\label{fig:log-scale-failures}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=.8\linewidth]{ft-reference-lines}
\caption{An annotated time series from the Financial Times which appeared on Twitter on November 12, 2020. This chart has annotations which show the decreasing $R_{eff}$, the effective rate of exponential spread, in successive peaks of the pandemic, with periods of $R_{eff}<1$ between surges in cases. These annotations assist the reader with drawing complex conclusions from the chart, but are difficult to automate and thus tend to be manually curated \citep{gubrudDrEricDingWeVe2020}.}
\label{fig:log-scale-ref-multiple}
\end{figure}

@@

\section{Summary and Discussion} \label{sec:summary}
Throughout the pandemic (thus far), visual representations of data have been an integral part of scientific communication. While not always optimal, these graphics have attempted to provide meaningful context, to encourage individual and collective action, and to help individuals grapple with the scale of the pandemic in cases, deaths, vaccinations, and interventions. As in any developing situation, choices made at some points in the pandemic did not always persist - many news outlets refined their charts and approach to the design of graphics over the course of the pandemic. This evolution of graphical forms has provided us with an opportunity to evaluate what worked and did not -- and why -- in a context of broad general interest.

@@

Archived data and code to create the figures in this paper can be found at \url{https://github.com/srvanderplas/covid-visualization-dssv}.

\section*{Archived Information}
This paper references many news articles which contain interactive graphics. In order to preserve these graphics (which depend on a great many external files), the authors have preserved many of the interactive graphics as self-contained html files \citep{news-file-archive}. 
The authors have also made extensive use of the Internet Wayback Machine, which aims to preserve the state of important web pages as they change over time.

@@

First ideas have been presented at DSSV 2020. We thank our colleagues, the reviewers, and the editors for valuable feedback and encouragement to complete and revise this manuscript.
